---
title: "Arithmetic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arithmetic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Enable densification for vignette examples
options(dbMatrix.allow_densify = TRUE)
```

```{r setup}
library(dbMatrix)
```


## dbMatrix arithmetic

`dbMatrix` objects support `Arith` and `Ops` operations. We will demonstrate how to perform arithmetic operations on `dbSparseMatrix` objects.

**Note:** Some operations with zero values are not yet supported with dbMatrix objects. In addition, certain arithmetic operations between `dbMatrix` objects are also not yet supported. We welcome user feedback and reporting issues on the [Github page ](https://github.com/dbverse-org/dbmatrix-r/).

### Create test data
Let's create a simple sparse matrix for demonstration:

```{r}
# Create a sparse matrix
set.seed(42)
dgc <- Matrix::rsparsematrix(100, 50, density = 0.1, rand.x = function(n) rpois(n, 5) + 1)
rownames(dgc) <- paste0("gene_", seq_len(100))
colnames(dgc) <- paste0("cell_", seq_len(50))

dplyr::glimpse(dgc)
```

The matrix contains 100 rows (genes) and 50 columns (cells). Like most single-cell RNA-seq data, the matrix is sparse.


### Create a dbMatrix object
Let's create a `dbSparseMatrix` object from the above `dgc` object.
```{r}
# Note: by default the constructor creates a dbMatrix object in-memory
con <- DBI::dbConnect(duckdb::duckdb(), ":memory:")

dbsm <- dbMatrix(
  value = dgc,
  con = con,
  name = "test_matrix",
  class = "dbSparseMatrix",
  overwrite = TRUE
)

# preview the object
dbsm
```

### Scalar Arithmetic
`dbMatrix` emulates scalar arithmetic in the `Matrix` package. 

Note: Addition or subtraction with non-zero addends on a `dbSparseMatrix` results in a `dbDenseMatrix`.
```{r} 
dbsm + 1

dbsm * 100
```

### Matrix Arithmetic
`dbMatrix` also supports matrix arithmetic for `dbMatrix` objects that are [conformable](https://en.wikipedia.org/wiki/Conformable_matrix).
```{r}
dbsm + dbsm
```

### Matrix Multiplication

#### Hadamard product
```{r}
dbsm * dbsm
```

#### Matrix product
TODO

### Cleanup
```{r}
DBI::dbDisconnect(con, shutdown = TRUE)
```

### Session Info
```{r}
sessionInfo()
```
